Post MCMCglmm analyses

### 2021-04-12 Analysis/06-post_mcmcglmm_analyses.Rmd
vector diff function

 * model 6: `model_phylo1_clade3` with one phylogenetic random effect (across the whole tree) and three clade residual effects.
 * model 7: `model_phylo3_clade3` with three phylogenetic random effect (one for each clade) and three clade residual effects.

```{r}
## Loading the correct models
load("../Data/Processed/model_list.rda")
model_phylo1_clade3 <- model_list[[4]]
model_phylo3_clade3 <- model_list[[6]]
```

@@

```{r, fig.height = 4, fig.width = 8}
par(mfrow = c(1,2))
plot.all.axes(uncentered_major_axes,
         col = c("grey",colour_vector),
         add = FALSE, main = "Major axes (un-centred)")
plot.all.axes(level_centred_major_axes,
         col = c("grey",colour_vector),
         add = FALSE, main = "Major axes (level-centred)")
```

### Phylo clade + clade model

```{r, warn = FALSE}
plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
     col = colour_vector[morphdat$clade],
     xlab = "PC1 (90.5%)",
     ylab = "PC1 (6.86%)")
plot.ellipses(model_phylo3_clade3, n = 100,
              centre = "none", add = TRUE, col = c("black", colour_vector, colour_vector))
```

This looks much more messier (and also probably not scaled correctly?)...

```{r}
plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
     col = colour_vector[morphdat$clade],
     xlab = "PC1 (90.5%)",
     ylab = "PC1 (6.86%)")
plot.ellipses(model_phylo3_clade3, n = 100,
              centre = "level", add = TRUE, col = c("black", colour_vector, colour_vector))
```

@@

```{r}
boxplot.proj.rej <- function(point_list, data, cols) {
    ## Get the classifier
    classifier <- data$clade    
    sort.clade.rej <- function(data, classifier) {
        return(cbind(data$rejection, classifier))
    }
    sort.clade.proj <- function(data, classifier) {
        return(cbind(data$projection, classifier))
    }
    ## Sort the data per clade
    rejections <- do.call(rbind, lapply(point_list, sort.clade.rej, classifier))
    projections <- do.call(rbind, lapply(point_list, sort.clade.proj, classifier))
    colnames(rejections)[1] <- colnames(projections)[1] <- "points"
    par(mfrow = c(2,1))
    boxplot(points ~ classifier, data = projections, col = cols, main = "Projections")
    boxplot(points ~ classifier, data = rejections, rejections, col = cols, main = "Rejections")
}

@@

TODO:

 - run the blob wise and tip wise models

>>>>

 * model 6: `model_phylo1_clade3` with one phylogenetic random effect (across the whole tree) and three clade residual effects.
 * model 7: `model_phylo4` with three phylogenetic random effect (one for each clade) and three clade residual effects.

```{r}
## Loading the correct models
load("../Data/Processed/model_list.rda")
model_phylo1_clade3 <- model_list[[4]]
model_phylo4 <- model_list[[7]]
```

@@

```{r}
plot.all.axes(uncentered_major_axes,
         col = c("grey",colour_vector),
         add = FALSE, main = "Major axes (un-centred)")
```

We can then compare these vectors per blobs.
We in order to facilitate the vectors comparisons, we will translate the groups vectors (clade ones) on the origin of the phylo vector.

```{r}
source("../Functions/vector.diff.R")
## Get all the angles and stuff
gull_results <- mapply(vector.diff, uncentered_major_axes$cladegulls, uncentered_major_axes$phylo)
plovers_results <- mapply(vector.diff, uncentered_major_axes$cladeplovers, uncentered_major_axes$phylo)
sandpipers_results <- mapply(vector.diff, uncentered_major_axes$cladesandpipers, uncentered_major_axes$phylo)

## Sort the results per type
angles <- cbind(gull_results["angle", ],
                plovers_results["angle", ],
                sandpipers_results["angle", ])
projections <- cbind(gull_results["projection", ],
                     plovers_results["projection", ],
                     sandpipers_results["projection", ])
rejections <- cbind(gull_results["rejection", ],
                    plovers_results["rejection", ],
                    sandpipers_results["rejection", ])
colnames(angles) <- colnames(projections) <- colnames(rejections) <- levels(morphdat$clade)
```

```{r, fig.height = 18, fig.heigh = 6}
## Plot the blob wise metrics
par(mfrow = c(3, 1))
boxplot(angles, col = colour_vector, ylab = "angle")
boxplot(projections, col = colour_vector, ylab = "projection (scaled)")
boxplot(rejections, col = colour_vector, ylab = "rejection (scaled)")
```

Same but with absolute values and angles modulo 90:

```{r, fig.height = 18, fig.heigh = 6}
## Plot the blob wise metrics
par(mfrow = c(3, 1))
boxplot(angles %% 90, col = colour_vector, ylab = "angle (modulo 90)")
boxplot(abs(projections), col = colour_vector, ylab = "projection (scaled - absolute)")
boxplot(abs(rejections), col = colour_vector, ylab = "rejection (scaled - absolute)")
```

### Phylo clade model

For this example, the MCMCglmm model has four random (phylo) effects (one for the whole phylo and one per clade) rather than just the one for the whole phylogeny.
The model seems to struggle converging and there are things off with the centring (they seem to be all centred no matter what) but that last bit should not be a problem for comparing the vectors since they are translated anyways.

```{r, warn = FALSE}
plot(morphdat[, c(1,2)], pch = 19, cex = 0.5,
     col = colour_vector[morphdat$clade],
     xlab = "PC1 (90.5%)",
     ylab = "PC1 (6.86%)")
plot.ellipses(model_phylo4, n = 100,
              centre = "none", add = TRUE, col = c("grey",colour_vector, "pink"))
```

The pink ellipse is representing the residuals.

And we can still do the whole axis stuff, same way as above:

```{r}
## Get the covariance matrices
covar_matrices2 <- get.covar(model_phylo4,
                            n = 100,
                            simplify = FALSE)
## Drop the residuals
covar_matrices2 <- covar_matrices2[-5, ]

## Get all the phylo main axes
uncentered_major_axes2 <- get.axes(covar_matrices2, centre = "none")
names(uncentered_major_axes2) <- c("phylo", "cladegulls", "cladeplovers", "cladesandpipers")

```

Here's what the axes look like without the ellipses and stuff:

```{r}
plot.all.axes(uncentered_major_axes2,
         col = c("grey",colour_vector),
         add = FALSE, main = "Major axes (centred)")
```

```{r}
source("../Functions/vector.diff.R")
## Get all the angles and stuff
gull_results <- mapply(vector.diff, uncentered_major_axes2$cladegulls, uncentered_major_axes2$phylo)
plovers_results <- mapply(vector.diff, uncentered_major_axes2$cladeplovers, uncentered_major_axes2$phylo)
sandpipers_results <- mapply(vector.diff, uncentered_major_axes2$cladesandpipers, uncentered_major_axes2$phylo)

## Sort the results per type
angles <- cbind(gull_results["angle", ],
                plovers_results["angle", ],
                sandpipers_results["angle", ])
projections <- cbind(gull_results["projection", ],
                     plovers_results["projection", ],
                     sandpipers_results["projection", ])
rejections <- cbind(gull_results["rejection", ],
                    plovers_results["rejection", ],
                    sandpipers_results["rejection", ])
colnames(angles) <- colnames(projections) <- colnames(rejections) <- levels(morphdat$clade)
```

```{r, fig.height = 18, fig.heigh = 6}
## Plot the blob wise metrics
par(mfrow = c(3, 1))
boxplot(angles, col = colour_vector, ylab = "angle")
boxplot(projections, col = colour_vector, ylab = "projection (scaled)")
boxplot(rejections, col = colour_vector, ylab = "rejection (scaled)")
```

```{r, fig.height = 18, fig.heigh = 6}
## Plot the blob wise metrics
par(mfrow = c(3, 1))
boxplot(angles %% 90, col = colour_vector, ylab = "angle (modulo 90)")
boxplot(abs(projections), col = colour_vector, ylab = "projection (scaled - absolute)")
boxplot(abs(rejections), col = colour_vector, ylab = "rejection (scaled - absolute)")
```

@@

```{r}
boxplot.proj.rej <- function(point_list, data, cols) {
    ## Get the classifier
    classifier <- data$clade    
    sort.clade.rej <- function(data, classifier) {
        return(data.frame("points" = data$rejection,
                          "classifier" = classifier))
    }
    sort.clade.proj <- function(data, classifier) {
        return(data.frame("points" = data$projection,
                          "classifier" = classifier))
    }
    ## Sort the data per clade
    rejections <- do.call(rbind, lapply(point_list, sort.clade.rej, classifier))
    projections <- do.call(rbind, lapply(point_list, sort.clade.proj, classifier))
    par(mfrow = c(2,1))
    boxplot(points ~ classifier, data = projections, col = cols, main = "Phylo projections\n(projected on the whole phylo effect)")
    boxplot(points ~ classifier, data = rejections, rejections, col = cols, main = "Phylo rejections")
}

@@

And now we can see the differences for the within clades (i.e. the elaboration/exploration within each clade).
The idea here is to distinguish between extra special taxa.
For example, in mammals the platypus and the naked mole rat are special, however, the platypus is actually not special among platypuses whereas the naked mole rat is special among rodents!

```{r}
## The within clades explo/elabo
gulls_results_uncentered <- lapply(uncentered_major_axes$cladegulls, lapply.projections, trait_space = morphdat[which(morphdat$clade == "gulls"), c(1,2,3)])
plover_results_uncentered <- lapply(uncentered_major_axes$cladeplovers, lapply.projections, trait_space = morphdat[which(morphdat$clade == "plovers"), c(1,2,3)])
sandpipers_results_uncentered <- lapply(uncentered_major_axes$cladesandpipers, lapply.projections, trait_space = morphdat[which(morphdat$clade == "sandpipers"), c(1,2,3)])

projection_list <- list("gulls"   = unlist(lapply(gulls_results_uncentered, `[[`, "projection")),
             "plovers" = unlist(lapply(plover_results_uncentered, `[[`, "projection")),
             "sandpipers" = unlist(lapply(sandpipers_results_uncentered, `[[`, "projection")))

rejection_list <- list("gulls"   = unlist(lapply(gulls_results_uncentered, `[[`, "rejection")),
             "plovers" = unlist(lapply(plover_results_uncentered, `[[`, "rejection")),
             "sandpipers" = unlist(lapply(sandpipers_results_uncentered, `[[`, "rejection")))
par(mfrow = c(2,1))
boxplot(projection_list, main = "projection within clades\n(projected on the clade specific effect)", col  = colour_vector)
boxplot(rejection_list, main = "rejection within clades", col  = colour_vector)
```

They seem a bit off as well.
Maybe it'll be worth centering their axis on their trait space centroid.

TODO.
